I am so grateful to my entire community{\emdash}in astronomy and beyond{\emdash}whose support, guidance, and friendship infuse the pages of this dissertation.
To my advisor, David (W.) Hogg: you have been a phenomenal teacher while always treating me as an equal colleague.
Thank you for teaching me that there's no such thing as thinking too deeply{\emdash}about approaching a research problem, making our field more equitable, or formatting {\LaTeX} (don't worry, this uses \verb|\frenchspacing|).
I never imagined a PhD would challenge not only my intellect but my worldview as much{\emdash}or be as fun{\emdash}as it has with you as an advisor.
To Jeremy Tinker: thank you for being the most supportive co-advisor, and for helping me become a real cosmologist.
To the rest of my thesis committee, thank you for your feedback year after year, and for having high expectations while always being kind: Michael Blanton, whose critical skepticism has made me a better scientist; David Grier, who asks the best big-picture questions; and Shy Genel, the most patient astronomer I know.
I have been incredibly lucky to work with so many engaged and thoughtful mentors and collaborators throughout my PhD (in the words of one interviewer, ``You've worked on so many projects with totally different sets of people{\emdash}how did that happen? Really, tell me!'').
Huge thank yous to Soledad Villar, Hans-Walter Rix, Marc Huertas-Company, Christina Eilers, Nesar Ramachandra, Fran\c{c}oise Lanusse, Zhongxu Zhai, Risa Wechsler, Joe DeRose, Giulio Fabbian, So Hattori, Austen Gabrielpillai, X. Prochaska, Alexie Leauthaud, Weichi Yao, Chris Lintott, and all of my other co-authors.
I have had the honor of mentoring Abby Williams, whose drive and humility have guided me{\emdash}in a fortuitous exchange{\emdash}through the better part of my PhD.

I can't imagine these six years without my NYU Physics cohort and friends. 
Thank you to Paul McNulty, to whom I'm forever grateful for helping my plants survive the pandemic, and me survive grad school; Lauren Altman, who has been my physics bud and venting outlet from undergrad to PhD; Austin McDowell, officemate extraordinaire; Chris Ick, gatherer of the Cosmic Voyagers; Jack Donahue, go-to CCPP back-channeler; and Iraj Eshghi, with whom I share so much despite being so different{\emdash}as evidenced by his inclusion in nearly all of these acknowledgement categories, and more.
I am so proud to have been part of the formation of our graduate organization, G-PHORCE, which wouldn't have happened without the devotion and charter-drafting skills of Marco Muzio, Christina Mondino, Paul McNulty, and Iraj Eshghi.
To all the NYU Physics graduate students, especially Suroor, Shahrzad, Ankit, Aditya, Trey, Cara, Clayton, Alexis, Ekapob, Chris, Marcus, Trey, Alex M., Michael, and Matt D. (both of you): witnessing your commitments to both excelling at your research and finishing every drop of colloquium wine will continue to inspire me far beyond 726.

When I attended last year's American Astronomical Society and European Astronomical Society meetings, I couldn't go long without running into a past mentor, conference friend, or astro-Twitter mutual.
To Chris Lovell, Dan Cunnama, and Pippa Cole, who embody the astro-friends-to-real-life-friends pipeline; Alex Gagliano, Jenny Calahan, and Hayley Beltz (solidarity squad forever); Gillian Beltz-Mohrmann and Lachlan Lancaster (how lucky I was to meet you at that Berlin summer school); Katie Breivik, Gaby Contardo, Adrian Price-Whelan, Lauren Anderson, and Viviana Acquaviva (CCA was good for more than just the cold brew on tap); and Emily Gilbert, Tom Rice, and Mike Walmsley, who have paved the way for me in the field: all of you are the reason I am proud to call myself an astronomer.
Joining Astrobites was by far one of the best things I've done in graduate school: it has shown me what it means to be part of a community that is passionate about a common goal while prioritizing care for each other. Special thanks to Astrobiters Briley, Mia, Jenny, Haley, Huei, Gourav, Nora, Michael H., Sabina, and Joanna.
Other organizations and groups have been deeply formative to my growth as a scientist and teacher: the ISEE Professional Development Program, the Kavli Summer Program in Astrophysics, the NYU Science Communication Workshop, the Flatiron CCA AstroData Group, and dotAstronomy. 
Thank you to my mentors and peers along each step of my astrophysics journey, from middle school Mathletes to Brown to REUs to Obs: Barry Bishop, Anton Krukowski, Jahmour Givans, K\"allan Berglund, Sheila Kannappan, Adrienne Erickcek,  Ari Maller, Alex Mannings, Romeel Dav\'e, and many more.
Finally, an Ode to Conference Friends: we might have only spent a week together and hardly spoken since{\emdash}but for that week we bonded over 7am hotel breakfasts, whispered during the conference talks, and discovered local dives in new cities.
Thank you for the companionship and making this field a community.

I am grateful to everyone in the NYU Physics department who supported me and my research over these years.
Thank you to the postdocs, whom I always looked up to, including Mehmet Alpaslan, Sarah Pearson, Abhishek Maniyar, Sjoert van Velzen, and Boris Leistedt.
I have been lucky to have the guidance of Professors Roman Scoccimarro, Anthony Pullen, Maryam Modjaz, and Andre Adler (the most dedicated and pure teacher-of-teachers).
I cannot thank the NYU Physics staff enough; the department would grind to a halt without each of them.
To Evette Ma, Lorelei DeMesa, Mulin Ding, Tim Johnston, Raj Sivanandarajah, Karven Lam, Stanley, and Juan: thank you for being constant lights and always-helpful resources through the graduate school slog.
To Laura, who has the coolest hair in the department: I'm so lucky to call you a friend.

I owe my union, GSOC--UAW Local 2110, the deepest gratitude for supporting me throughout graduate school as both a worker and an activist. 
GSOC has shown me the meaning of collective action{\emdash}and, after a hard-won fight, of getting the goods.
To my comrades and friends Daniel, Caroline, Bhumika, and Iraj (seize! the! memes!); Sandy, Ellis, Dylan, Meg, Yuwei, Elsy, Zach, Kate B., Colin, Virgilio, Grace, Peter, Naveen, Paul, Chris N., Jennifer; and the rest of AWDU and GSOC: I am forever awed by your organizing prowess and commitment to our shared ideals.
You have made academia a better place, and me a better person. 
(And, incidentally, a better physicist{\emdash}it turns out organizing skills are really useful for research.)
Solidarity forever.

This work would not be possible without my friends and family bringing the ``life'' to work--life balance.
Alexa VanHattum: I wish everyone were so lucky to have such a friendship as ours.
I couldn't have gotten through undergrad, adult life, or job season without you being just a room, phone call, or park-walk away, and I can't wait to see where our near-parallel lives take us.
To the Castle: thank you for showing me the meaning of living in community and teaching me that time is best counted in kitchen squats, stoop sits, and backyard jams.
You'll be the first to know if I ever have an answer to ``Have you found the dark matter yet?''
Stoddard and friends, you transformed New York life into a constant string of adventures; thank you for being a second home.
To my book club, thank you for being an unrivaled constant through my NYC years, and to my writing group, for providing me the space to create strange worlds and helping me understand what is important in this world.
To Lyra, Kotik, and Rua: thank you for teaching me the importance of playing, being outside, and screaming.
My friends are the many suns around which my life revolves: to Sammy, Caitlin, Paul Elliott, Tiff, Mickey, Nicole, Kailani, Daniella, Maryann, Noel, Lila, Alison, Rachel, Alex, Gabby, Cali, Melissa, David, Ruthie, and Ariella, I am dizzy with gratification for your friendship.
And to Zeke, who inspires me to live life with arms wide open.

I am infinitely grateful to my family for supporting my path in academia, whether it takes me to the other side of the globe or right back home (!).
Thank you to my mother, Jill, whose astute guidance and sharp wit get me through each day (and whose sharp editing helps get my grant proposals funded); my father, Richard, whose commitment to having fun{\emdash}from traveling the world to backyard mini-golf{\emdash}has shaped my life choices; my brother, Ry, who has taught me the importance of pursuing both one's passion and the best iced oat milk latte; and my sis, Maya, whose depth of care and creativity still floors me (thanks to you I have skinned the cow!).
To my extended family, including Grandma Sylvia, Papa Stan, Bev, Shireen, and all the Storeys and Fishers: thank you for always being my most enthusiastic supporters (and for a memorable publishing name).

Finally, to the readers who have made it to the end of these acknowledgements: thank you!
You likely understand, as I have learned, that the most important part of this work is the people.